From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments

### 2015-09-14 write_up.tex
working from tex now

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\date{}

\begin{document}

Causal inference -- the task of uncovering regulatory relationships
between components of biomolecular pathways and networks - is a primary
goal of omics datasets in molecular biology studies. Statistical
associations can reveal an alluring set of putative causal interactions,
but when are such associations significant and reliable? And when can
they be interpreted as causal? Our goal below is to provide suggestions
for causal inference in large scale experiments, such as those resulting
from high throughput omics technologies. We describe the pitfalls of
large datasets, and suggest methods to reliably find causal associations
despite these pitfalls.

High-throughput omics technologies enable large-scale experiments, with
simultaneous measurements of multiple cellular components in a sample.
The huge amount of data generated paradoxically make the task of causal
inference much more difficult, because, when many associations are
examined, some will appear significant just by chance. In other words,
large-scale experiments are more prone to hiding the true signal and
generating spurious associations, leading to increased false positive
conclusions that causal relationships are present. This problem can be
addressed by improving experimental design and/or refining the
biological question. Below, we describe the process of elucidating
statistical associations from data and suggest practical approaches for
analysis of large scale datasets.

I. Statistical inference of causal relationships

@@

\begin{enumerate}
\def\labelenumi{\Roman{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Inference with large numbers of variables
\end{enumerate}

@@

Fig1. In the MAPK signaling pathway, Raf regulates Mek, which regulates
Erk. Top; the lines represent association - the activity of the kinases
on this pathway are all correlated. Bottom: the lines represent
conditional dependence, there is no conditional dependence edge from Raf
to Erk because given you know Mek, knowing Raf tells you nothing about
Erk. They are conditionally independent.

@@

Fig2. As you increase the number of features, you get more high scoring
spurious correlations.

@@

Fig3. Increasing the number of measured features means increasing the
false positive detection of conditional dependence.

@@

\begin{enumerate}
\def\labelenumi{\Roman{enumi}.}
\setcounter{enumi}{2}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Inferring causality from omics experiments
\end{enumerate}

The problems outlined in section II paint a grim picture for causal
learning in large datasets. Fortunately, these can be overcome, and
effective causal learning can be a reality for large scale datasets. We
proffer the approaches listed in the Causality Inference in Omics tools
below.

Refine the biological problem, thus limiting the number of features. If
the broader biological system is well-studied, it may be possible design
an experiment that focuses on a specific part of the system of interest,
then ask more specific questions of the data, such as whether a
particular edge or pathway is present. The more specific the question,
the less data is needed overall to make solid statistical inferences.

Measure more samples. If feasible, high throughput measurements that
measure many samples (not just many features) provide the statistical
power to tell true associations from spurious associations. This
intuition motivates the use technologies that have lower throughput but
enable measurement of more samples (or gather more data points per
sample), for causal modeling. An example is single cell mass cytometry,
where many thousands of cells per sample provide ample statistical
power.

Use prior knowledge, not just from experts but also from noisier
sources, to improve the process of searching for conditional
independence. The MAPK pathway for instance, mentioned earlier, is well
established and canonical. Assuming it, and other well established
connections a priori can help reduce the number of possible
associations, and enable more effective use of data. Knowledge about
canonical causal relationships can be sourced from pathway databases
such as KEGG. Another example is contextual information, such as spatial
relations between phosphoproteins in the cell -- causal inference
algorithms can be formulated to weigh evidence of conditional dependence
differently depending on whether phosphoproteins are from the same or
different spatial compartments. Employing prior knowledge that provides
evidence for the regulatory network reduces the number of connections
and causal arcs under consideration, which allows the available data to
make reliable statistical inferences.

Employ targeted interventions selectively. Targeted interventions
perturb individual components of the system. An example is small
molecule inhibitors which block the causal influence of a specific
phoshoprotein on downstream components. Since it is usually prohibitive
to perturb every target component of a system, it is strategic to
prioritize application of targeted interventions to parts of the system
that have the most potential for new discoveries. Selection of these
perturbation targets can be based on prior knowledge - e.g.~knowledge of
which components are crucial players in the system of interest - or they
can be applied iteratively, after an initial statistical analysis has
revealed areas of the network in which causal inferences are not
possible based on existing data. For instance, a resulting model graph
with undirected edges can be inspected to reveal which nodes have
potential to reveal the most causality if perturbed.

Consider broad-scale interventions. Traditionally, experiments consider
a biological response to a particular stimulus, possibly in the context
of an inhibitor. In contrast, broad-scale interventions sacrifice
specificity to perturb many features in the system simultaneously. The
advantage of this approach is that it can enable elucidation of
causality across the entire system. Just as one intervention was
sufficient to infer causality between 3 proteins Raf, Mek, and Erk,
perturbing many things at once creates cascades of causal direction
orientation across the broader network. This includes varying
experimental conditions to activate multiple pathways. For example,
signals from endocrine, paracrine, and autocrine ligands elicit various
signaling responses in hepatocytes, thus interventions that cover this
range of signals gives the best picture of the broader causal network of
hepatocyte signaling. Similarly, interventions that go beyond
receptor-level and perturb multiple components of the system bring
cascading causal direct orientation deeper into the network. In contrast
to targeted interventions, while it is difficult to know in advance what
information will provided by broad-scale interventions, they provide
more causal inference bang for your intervention buck.

The Causality in Omics tool list above provides impactful approaches
that can drastically improve causal inference from omics datasets, by
constraining the inference task, and thus allowing for accurate
statistical inferences. For instance, the task of assessing which of all
the possible KEGG pathways is present in a dataset will be far less
error-prone than the task of assessing which of all possible
combinations of my measured features might form a biological pathway.

How should the tools listed be used? They are most powerful when used in
combination, and in fact the lines between them are somewhat arbitrary
and frequently blurred. For instance, using tool \#1 and tool \#2 in
concert can be thought of as reducing the breadth and increasing the
depth of the investigation. Tools \#4 and \#5 call for use of
interventions, but this task itself is complicated by measuring many
things, since we have more features to expose to intervention. . Tool
\#3, prior biological knowledge, can be used to prioritize what to
target with that limited set of interventions.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% This is a (brief) model paper using the achemso class
%% The document class accepts keyval options, which should include
%% the target journal and optionally the manuscript type.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[journal=jacsat,manuscript=article]{achemso}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional packages needed here.  Only include packages
%% which are essential, to avoid problems later. Do NOT use any
%% packages which require e-TeX (for example etoolbox): the e-TeX
%% extensions are not currently available on the ACS conversion
%% servers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
\usepackage[T1]{fontenc}       % Use modern font encodings

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If issues arise when submitting your manuscript, you may want to
%% un-comment the next line.  This provides information on the
%% version of every file you have used.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%\listfiles

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Place any additional macros here.  Please use \newcommand* where
%% possible, and avoid layout-changing macros (which are not used
%% when typesetting).
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Meta-data block
%% ---------------
%% Each author should be given as a separate \author command.
%%
%% Corresponding authors should have an e-mail given after the author
%% name as an \email command. Phone and fax numbers can be given
%% using \phone and \fax, respectively; this information is optional.
%%
%% The affiliation of authors is given after the authors; each
%% \affiliation command applies to all preceding authors not already
%% assigned an affiliation.
%%
%% The affiliation takes an option argument for the short name.  This
%% will typically be something like "University of Somewhere".
%%
%% The \altaffiliation macro should be used for new address, etc.
%% On the other hand, \alsoaffiliation is used on a per author basis
%% when authors are associated with multiple institutions.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Robert O. Ness}
\email{nessr@purdue.edu}
\affiliation[Purdue University]{Department of Statistics, Purdue University, West Lafayette}
\author{Karen Sachs}
\affiliation[Stanford University]{School of Medicine, Stanford University, Palo Alto}
\email{karens1@stanford.edu}
\author{Olga Vitek}
\email{o.vitek@neu.edu}
\affiliation[Northeastern University]{College of Science, College of Computer and Information Science, Northeastern University, Palo Alto}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {From Correlation to Causality: Statistical approaches to learning regulatory relationships from Large-scale Omics Experiments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{IR,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The manuscript does not need to include \maketitle, which is
%% executed automatically.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Start the main part of the manuscript here.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Causal inference -- the task of uncovering regulatory relationships
between components of biomolecular pathways and networks - is a primary
goal of omics datasets in molecular biology studies. Statistical
associations can reveal an alluring set of putative causal interactions,
but when are such associations significant and reliable? And when can
they be interpreted as causal? Our goal below is to provide suggestions
for causal inference in large scale experiments, such as those resulting
from high throughput omics technologies. We describe the pitfalls of
large datasets, and suggest methods to reliably find causal associations
despite these pitfalls.

High-throughput omics technologies enable large-scale experiments, with
simultaneous measurements of multiple cellular components in a sample.
The huge amount of data generated paradoxically make the task of causal
inference much more difficult, because, when many associations are
examined, some will appear significant just by chance. In other words,
large-scale experiments are more prone to hiding the true signal and
generating spurious associations, leading to increased false positive
conclusions that causal relationships are present. This problem can be
addressed by improving experimental design and/or refining the
biological question. Below, we describe the process of elucidating
statistical associations from data and suggest practical approaches for
analysis of large scale datasets.
\section{Discussion}
\subsection{I. Statistical inference of causal relationships}

@@

\subsection{II. Inference with large numbers of variables}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.5\textwidth]{figs/mapk.jpg}}
\caption{In the MAPK signaling pathway, Raf regulates Mek, which regulates
Erk. Top; the lines represent association - the activity of the kinases
on this pathway are all correlated. Bottom: the lines represent
conditional dependence, there is no conditional dependence edge from Raf
to Erk because given you know Mek, knowing Raf tells you nothing about
Erk. They are conditionally independent.}
\label{mapk}
\end{figure}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_corr.png}}
\caption{As you increase the number of features, you get more high scoring
spurious correlations.}
\label{spur_corr}
\end{figure}

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.5\textwidth]{figs/spurious_dep.png}}
\caption{Fig3. Increasing the number of measured features means increasing the
false positive detection of conditional dependence.}
\label{spur_dep}
\end{figure}

@@

\subsection{III. Inferring causality from omics experiments}

The problems outlined in section II paint a grim picture for causal
learning in large datasets. Fortunately, these can be overcome, and
effective causal learning can be a reality for large scale datasets. We
proffer the approaches listed in \textit{The Causality Inference in Omics Tool List}
below.

\begin{enumerate}
\item \textit{Refine the biological problem}, thus limiting the number of features. If
the broader biological system is well-studied, it may be possible design
an experiment that focuses on a specific part of the system of interest,
then ask more specific questions of the data, such as whether a
particular edge or pathway is present. The more specific the question,
the less data is needed overall to make solid statistical inferences.
\item \textit{Measure more samples}. If feasible, high throughput measurements that
measure many samples (not just many features) provide the statistical
power to tell true associations from spurious associations. This
intuition motivates the use technologies that have lower throughput but
enable measurement of more samples (or gather more data points per
sample), for causal modeling. An example is single cell mass cytometry,
where many thousands of cells per sample provide ample statistical
power.
\item \textit{Use prior knowledge}, not just from experts but also from noisier
sources, to improve the process of searching for conditional
independence. The MAPK pathway for instance, mentioned earlier, is well
established and canonical. Assuming it, and other well established
connections a priori can help reduce the number of possible
associations, and enable more effective use of data. Knowledge about
canonical causal relationships can be sourced from pathway databases
such as KEGG. Another example is contextual information, such as spatial
relations between phosphoproteins in the cell -- causal inference
algorithms can be formulated to weigh evidence of conditional dependence
differently depending on whether phosphoproteins are from the same or
different spatial compartments. Employing prior knowledge that provides
evidence for the regulatory network reduces the number of connections
and causal arcs under consideration, which allows the available data to
make reliable statistical inferences.
\item \textit{Employ targeted interventions selectively}. Targeted interventions
perturb individual components of the system. An example is small
molecule inhibitors which block the causal influence of a specific
phoshoprotein on downstream components. Since it is usually prohibitive
to perturb every target component of a system, it is strategic to
prioritize application of targeted interventions to parts of the system
that have the most potential for new discoveries. Selection of these
perturbation targets can be based on prior knowledge - e.g.~knowledge of
which components are crucial players in the system of interest - or they
can be applied iteratively, after an initial statistical analysis has
revealed areas of the network in which causal inferences are not
possible based on existing data. For instance, a resulting model graph
with undirected edges can be inspected to reveal which nodes have
potential to reveal the most causality if perturbed.
\item \textit{Consider broad-scale interventions}. Traditionally, experiments consider
a biological response to a particular stimulus, possibly in the context
of an inhibitor. In contrast, broad-scale interventions sacrifice
specificity to perturb many features in the system simultaneously. The
advantage of this approach is that it can enable elucidation of
causality across the entire system. Just as one intervention was
sufficient to infer causality between 3 proteins Raf, Mek, and Erk,
perturbing many things at once creates cascades of causal direction
orientation across the broader network. This includes varying
experimental conditions to activate multiple pathways. For example,
signals from endocrine, paracrine, and autocrine ligands elicit various
signaling responses in hepatocytes, thus interventions that cover this
range of signals gives the best picture of the broader causal network of
hepatocyte signaling. Similarly, interventions that go beyond
receptor-level and perturb multiple components of the system bring
cascading causal direct orientation deeper into the network. In contrast
to targeted interventions, while it is difficult to know in advance what
information will provided by broad-scale interventions, they provide
more causal inference bang for your intervention buck.
\end{enumerate}

\textit{The Causality in Omics Tool List} above provides impactful approaches
that can drastically improve causal inference from omics datasets, by
constraining the inference task, and thus allowing for accurate
statistical inferences. For instance, the task of assessing which of all
the possible KEGG pathways is present in a dataset will be far less
error-prone than the task of assessing which of all possible
combinations of my measured features might form a biological pathway.

How should the tools listed be used? They are most powerful when used in
combination, and in fact the lines between them are somewhat arbitrary
and frequently blurred. For instance, using tool \#1 and tool \#2 in
concert can be thought of as reducing the breadth and increasing the
depth of the investigation. Tools \#4 and \#5 call for use of
interventions, but this task itself is complicated by measuring many
things, since we have more features to expose to intervention. Tool
\#3, prior biological knowledge, can be used to prioritize what to
target with that limited set of interventions.
